A comparison of design-based and model-based approaches for finite population spatial data.

### 2022-04-13 inst/manuscript/manuscript.tex
section 3.1 updates

CHANGE LOGNORMAL VERBAGE TO SKEWED -- look for DRE acronym

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (and there were no
replicates like for ZMMI). The ZMMI and Hg ppb data are shown spatially
and as a histogram in Figure \ref{fig:zmmi}. The ZMMI data are
relatively symmetric, and the Hg ppb data are skewed. Also in Figure
\ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb.
The spatial semivariogram quantifies the average squared differences
(semivariance) of responses whose separation (distance) falls within
some distance class. The spatial semivariance is closely related to the
spatial covariance, and spatial semivariograms are often used to gauge
the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to
have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as
the semivariance increases steadily with distance (meaning that
observations nearby one another tend to be more similar than
observations far apart from one another).

@@

We studied performance of the four sampling-inference combinations by
selecting 2000 random IRS and GRTS samples of size \(n = 50\),
\(n = 100\), and \(n = 200\) from the realized ZMMI and Hg ppb
populations and then analyzing the samples using MB and DB inference. We
used the same evaluation metrics as for the simulated data: mean bias,
RMSE, and interval coverage. Mean bias is taken as the average deviation
between each sample's estimated (or predicted) mean and the population
mean (of ZMMI or Hg ppb):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)\), where \(i\)
indexes simulation trials and \(\mu\) is the population mean.
Root-mean-squared error is taken as the square root of the average
squared deviation between each sample's estimated (or predicted) mean
and its population mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}\).
Interval coverage is taken as the proportion of simulation trials where
the population mean is contained in its 95\% confidence (or prediction)
interval. These intervals are constructed using the normal distribution.

@@

The mean bias was nearly zero for all four sampling-inference
combinations in all 36 scenarios, so we omit a more detailed summary of
those results here. Tables for mean bias in all 36 simulation scenarios
are provided in the supporting information.

Fig. \ref{fig:rmspe_eff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-inference combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:rmspe_eff}, ``Prop DE: 0'' row), the four sampling-inference
combinations have approximately equal rMS(P)E. In these scenarios, using
the GRTS algorithm or a model-based analysis does not increase
efficiency compared to IRS-Design. When there is spatial covariance
(Fig. \ref{fig:rmspe_eff}, ``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows),
GRTS-Model tends to have the lowest rMS(P)E, followed by GRTS-Design,
IRS-Model, and finally IRS-Design, though the difference in relative
rMS(P)E among GRTS-Model, GRTS-Design, and IRS-Model is relatively
small. As the strength of spatial covariance increases, the gap in
rMS(P)E between IRS-Design and the other sampling-inference combinations
widens. Finally we note that when there is spatial covariance, IRS-Model
has a much lower rMS(P)E than IRS-Design, suggesting that the poor
design properties of IRS are largely mitigated by the model-based
analysis. These rMS(P)E conclusions are similar to those observed in the
grid location layout, so we omit a grid location layout figure here.
Tables for rMS(P)E in all 36 simulation scenarios are provided in the
supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E in the simulation study for the four sampling-inference combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:rmspe_eff}
\end{figure}

Fig. \ref{fig:mse_eff} shows the relative MStdE of the four
sampling-inference combinations using the random location layout with
``IRS-Design'' as the baseline. The relative MStdE is defined as
\begin{equation*}
\frac{\text{MStdE of sampling-inference combination}}{\text{MStdE of IRS-Design}},
\end{equation*} Many general takeaways regarding MStdE are similar to
general takeaways regarding rMS(P)E: there seems to be no benefit to
using IRS, even when there is no spatial covariance; as the strength of
spatial covariance increases, the gap in MStdE between IRS-Design and
the other sampling-inference combinations widens; and IRS-Model
outperforms IRS-Design by a noticeable margin. These fact that the
rMS(P)E and MStdE findings are similar is not particularly surprising
because the mean bias for all sampling-inference combinations was nearly
zero, thus rMS(P)E is driven by the standard error of the estimators
(design-based) or predictors (model-based). We do note that between
GRTS-Design and GRTS-Model, GRTS-Design had lower MStdE when there was
no spatial covariance or a medium amount of spatial covariance (Fig.
\ref{fig:mse_eff}, ``Prop DE: 0'' and ``Prop DE: 0.5'' rows), and
GRTS-Model had lower MStdE when there was a high amount of spatial
covariance (Fig. \ref{fig:mse_eff}, ``Prop DE: 0.9'' row). These MStdE
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for MStdE in all
36 simulation scenarios are provided in the supporting information.

Fig. \ref{fig:figconf} shows the 95\% interval coverage for each of the
four sampling-inference combinations in the random location layout.
Within each scenario, the sampling-inference combinations tend to have
fairly similar interval coverage, though when \(n = 50\) or \(n = 100\),
GRTS-Design coverage is usually a few percentage points lower than the
other combinations. Coverage in the normal response scenarios was
usually near 95\%, while coverage in the lognormal response scenarios
usually varied from 90\% to 95\% but increased with the sample size. At
a sample size of 200, all four sampling-inference combinations had
approximately 95\% interval coverage in both response scenarios for all
dependent error proportions. These interval coverage conclusions are
similar to those observed in the grid location layout, so we omit a grid
location layout figure here. Tables for interval coverage in all 36
simulation scenarios are provided in the supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/coverage.jpeg}
  \caption{Interval coverage in the simulation study for the four sampling-inference combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line represents 95\% coverage.}
  \label{fig:figconf}
\end{figure}

\hypertarget{sec:r_app}{%
\subsection{National Lakes Assessment DAta}\label{sec:r_app}}

USE MERCURY UNITS

Fig. \ref{fig:merc} shows a map and histogram of mercury concentration
in all 986 NLA lakes. The map shows mercury concentration exhibits some
spatial patterning, with high mercury concentrations in the northeast
and north central United States. The histogram shows that mercury
concentration is right-skewed, with most lakes having a low value of
mercury concentration but a few having a much higher concentration. Fig.
\ref{fig:merc} also shows mercury concentration's empirical
semivariogram. The empirical semivariogram can be used as a tool to
visualize spatial dependence. It quantifies the mean of the halved
squared differences (semivariance) among all pairs of mercury
concentrations at different distances apart. When a process has spatial
covariance (exhibits spatial dependence), the mean semivariance tends to
be smaller at small distances and larger at large distances. The
empirical semivariogram in Fig. \ref{fig:merc} suggests that mercury
concentration exhibits spatial dependence. Lastly we note that the true
mean mercury concentration in the 986 NLA lakes is 103.2 ng / g.

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and constructed 95\% confidence (design-based) and 95\% (model-based)
prediction intervals. For the model-based analyses, the exponential
covariance was used. Table \ref{tab:appliedtab} shows the results from
these analyses. Though we should not generalize these results to other
samples from this population, we do mention a few findings. First,
IRS-Design has the largest standard error. Second, compared to
IRS-Design and IRS-Model, GRTS-Design and GRTS-Model are much closer to
the true mean mercury concentration (have bias closer to zero) and have
much lower standard errors (more precise intervals). Third, GRTS-Model
has the least amount of bias and the lowest standard error (most precise
interval). Finally, we note that for all sampling-inference
combinations, the true mean mercury concentration (103.2 ng / g) is
within the bounds of the combination's 95\% interval.

\begin{table}[ht]
\centering
\begin{tabular}{lrrrrr}
  \hline
Approach & True Mean & Est/Pred & SE & 95\% LB & 95\% UB \\ 
  \hline
IRS-Design & 103.2 & 112.7 & 8.8 & 95.4 & 129.9 \\ 
  IRS-Model & 103.2 & 110.5 & 7.9 & 95.0 & 125.9 \\ 
  GRTS-Design & 103.2 & 101.8 & 6.1 & 89.8 & 113.7 \\ 
  GRTS-Model & 103.2 & 102.3 & 5.9 & 90.8 & 113.9 \\ 
   \hline
\end{tabular}
\caption{\label{tab:appliedtab} For each sampling-inference combination (Approach), the true mean mercury concentration (True Mean), estimates/predictions (Est/Pred), standard errors (SE), lower 95\% interval bounds (95\% LB), and upper 95\% interval bounds (95\% UB) for mean mercury concentration computed using a sample of 100 lakes in the NLA data.} 
\end{table}

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach relies on
random sampling to estimate population parameters. The model-based
approach relies on distributional assumptions to predict realized values
of a stochastic process. Though the model-based approach does not rely
on random sampling, it can still be beneficial as a way to guard against
preferential sampling. While the design-based and model-based approaches
have often been compared in the literature from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling and the
design-based, local neighborhood variance estimator. Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, independent of the
analysis approach, we found no reason to prefer IRS over GRTS when
sampling spatial data -- GRTS-Design and GRTS-Model generally had
similar rMS(P)E as their IRS counterparts when there was no spatial
covariance and lower rMS(P)E than their IRS counterparts when there was
spatial covariance. Second, the sampling decision (IRS vs GRTS) is most
important when using a design-based analysis. Though GRTS-Model still
had lower rMS(P)E than IRS-Model, the model-based analysis mitigated
most of the rMS(P)E inefficiencies that result from the IRS samples
lacking spatial balance. Third, as the strength of spatial covariance
increases, the gap in rMS(P)E and MStdE between IRS-Design and the other
sampling-inference combinations also increases, likely because
IRS-Design is the only combination that ignores spatial locations in
sampling and analysis. Fourth and finally, when the response was normal,
interval coverage for all sampling-inference combinations was usually
close to 95\% for all sample sizes; when the response was lognormal,
interval coverage for all sampling-inference combinations was usually
between 90\% and 95\% and closest to 95\% when \(n = 200\).

>>>>

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (and there were no
replicates like for ZMMI). The ZMMI and Hg ppb data are shown as spatial
maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to
be highest near the coasts, lowest in the Central United States, higher
near the coasts, are relatively symmetric, and have a mean of 55.05. The
Hg ppb data tend to be highest in the Northeastern United States, lowest
elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure
\ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb.
The spatial semivariogram quantifies the the halved average squared
differences (semivariance) of responses whose separation (distance)
falls within some distance class. The spatial semivariance is closely
related to the spatial covariance, and spatial semivariograms are often
used to gauge the strength of spatial dependence in data. Both ZMMI and
Hg ppb seem to have moderately strong spatial dependence (Figure
\ref{fig:zmmi}), as the semivariance increases steadily with distance
(meaning that observations nearby one another tend to be more similar
than observations far apart from one another).

@@

We studied performance of the four sampling-inference combinations by
selecting 2000 random IRS and GRTS samples of size \(n = 50\),
\(n = 100\), and \(n = 200\) from the realized ZMMI and Hg ppb
populations and then analyzing the samples using MB and DB inference. In
total, there were six separate scenarios (two responses and three sample
sizes). We used the same evaluation metrics as for the simulated data:
mean bias, RMSE, and interval coverage. Mean bias is taken as the
average deviation between each sample's estimated (or predicted) mean
and the population mean (of ZMMI or Hg ppb):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)\), where \(i\)
indexes simulation trials and \(\mu\) is the population mean.
Root-mean-squared error is taken as the square root of the average
squared deviation between each sample's estimated (or predicted) mean
and its population mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}\).
Interval coverage is taken as the proportion of simulation trials where
the population mean is contained in its 95\% confidence (or prediction)
interval. These intervals are constructed using the normal distribution.

@@

Mean bias was nearly zero for all four sampling-inference combinations
in all 36 scenarios, so we omit a more detailed summary of those results
here. Tables for mean bias in all 36 simulation scenarios are provided
in the supporting information.

We define the relative RMSE as a ratio with numerator given by the RMSE
for a sampling-inference combination and the denominator given by the
RMSE for SRS-DB. Relative RMSEs for the random location layout are
provided in Fig. \ref{fig:rmspe_eff}. When there is no spatial
covariance (Fig. \ref{fig:rmspe_eff}, ``DRE\%: 0\%''), the four
sampling-inference combinations have approximately equal RMSE. In these
scenarios, using GRTS sampling or model-based inference does not
generally increase efficiency compared to SRS-DB. When there is spatial
covariance (Fig. \ref{fig:rmspe_eff}, ``DRE\%: 0\%'' and ``DRE\%:
50\%''), GRTS-MB tends to have the lowest RMSE, followed by GRTS-DB,
SRS-MB, and finally SRS-DB, though the difference in relative RMSE among
GRTS-MB, GRTS-DB, and SRS-MB is small. As the strength of spatial
covariance increases, the gap in RMSE between SRS-DB and the other
sampling-inference combinations widens. Finally we note that when there
is spatial covariance, SRS-MB has a much lower RMSE than SRS-DB,
suggesting that the lack of efficiency from SRS is largely mitigated by
model-based inference. These RMSE conclusions are similar to those
observed in the grid location layout, so we omit a figure and discussion
regarding the grid location layout here. Tables for RMSE in all 36
simulation scenarios are provided in the supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes}
  \label{fig:rmspe_eff}
\end{figure}

Fig. \ref{fig:figconf} shows 95\% interval coverage for each of the four
sampling-inference combinations in the random location layout. Within
each simulation scenario, all sampling-inference combinations tend to
have fairly similar interval coverage, though when \(n = 50\) or
\(n = 100\), GRTS-DB coverage is usually a few percentage points lower
than the other combinations, which suggests that the local neighborhood
variance estimate may be slightly too small for small \(n\). Coverage in
the normal response scenarios was usually near 95\%, while coverage in
the skewed response scenarios usually varied from 90\% to 95\% but
increased with the sample size. At a sample size of 200, all four
sampling-inference combinations had approximately 95\% interval coverage
in both response scenarios for all dependent error proportions. These
interval coverage conclusions are similar to those observed in the grid
location layout, so we omit a figure and discussion regarding the grid
location layout here. Tables for interval coverage in all 36 simulation
scenarios are provided in the supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/coverage.jpeg}
  \caption{Interval coverage in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid black lines separate the sample sizes and the dashed black lines represent 95\% coverage.}
  \label{fig:figconf}
\end{figure}

\hypertarget{sec:r_app}{%
\subsection{National Lakes Assessment (Real) Data}\label{sec:r_app}}

Mean bias was nearly zero for all four sampling-inference combinations
in all six scenarios, so we omit a more detailed summary of those
results here. Tables for mean bias in all six simulations scenarios are
provided in the supporting information.

Talk about RMSE

Talk about interval coverage -- mention how the model-based assumptions
of coverage are not really satisfied here, but they can be thought of as
``approximate''.

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach relies on
random sampling to estimate population parameters. The model-based
approach relies on distributional assumptions to predict realized values
of a stochastic process. Though the model-based approach does not rely
on random sampling, it can still be beneficial as a way to guard against
preferential sampling. While the design-based and model-based approaches
have often been compared in the literature from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling and the
design-based, local neighborhood variance estimator. Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, independent of the
analysis approach, we found no reason to prefer IRS over GRTS when
sampling spatial data -- GRTS-DB and GRTS-MB generally had similar RMSE
as their IRS counterparts when there was no spatial covariance and lower
RMSE than their IRS counterparts when there was spatial covariance.
Second, the sampling decision (IRS vs GRTS) is most important when using
a design-based analysis. Though GRTS-MB still had lower RMSE than
SRS-MB, the model-based analysis mitigated most of the RMSE
inefficiencies that result from the IRS samples lacking spatial balance.
Third, as the strength of spatial covariance increases, the gap in RMSE
and MStdE between SRS-DB and the other sampling-inference combinations
also increases, likely because SRS-DB is the only combination that
ignores spatial locations in sampling and analysis. Fourth and finally,
when the response was normal, interval coverage for all
sampling-inference combinations was usually close to 95\% for all sample
sizes; when the response was lognormal, interval coverage for all
sampling-inference combinations was usually between 90\% and 95\% and
closest to 95\% when \(n = 200\).
